Which groups are at highest risk for low uptake of flu vaccine?

### 2023-05-05 code/analysis_code/exploratory_analysis.qmd
More edits v2

# Setup
```{r, warning=FALSE, message=FALSE}
library(here) #for data loading/saving
library(dplyr)
library(ggplot2)
library(haven)
library(gridExtra) # To combine plots
library(gtsummary) # For producing table 1
#library("kableExtra") # For producing a nice table 1
```

@@

# Data exploration through figures

```{r}
getwd()
```

@@

```{r, warning=FALSE, message=FALSE}
library(table1)
(table1(~ factor(sex) + age + factor(race) + factor(hispanic) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
library(table1)
(table1(~ factor(education) + factor(fam_income) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
library(table1)
(table1(~ factor(employ_status) + factor(paid_sick) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
library(table1)
(table1(~ factor(health_ins_status) + factor(med_access) + factor(med_care_delayed_cost) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
library(table1)
(table1(~ factor(health_status) + factor(bmi) + factor(asthma) + factor(cancer) + factor(chd) + factor(alcohol_status) + factor(smoking_status) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
library(table1)
(table1(~factor(worry_freq) + factor(dep_freq) | flu_vaccine, data=mydata2))

@@

```{r}
tbl_summary(mydata2)
```

```{r}
a <- mydata2 %>% select(age, sex)
b <- tbl_summary(a)
```

## Full table for demogrophics characteristics
```{r, warning=FALSE, message=FALSE}
library(table1)
table_dems <- (table1(~ sex + age + race + hispanic + education + fam_income + employ_status + paid_sick + health_ins_status + med_access + med_care_delayed_cost + health_status + bmi + asthma + cancer + chd + alcohol_status + smoking_status| flu_vaccine, data=mydata2))

@@

@@

```{r}
install.packages("kableExtra", dependencies = TRUE)
```

@@

>>>>

# Setup
```{r, warning=FALSE, message=FALSE}
library(here) #for data loading/saving
library(dplyr)
library(ggplot2)
library(haven)
library(gridExtra) # To combine plots
library(gtsummary) # For producing table 1
library("kableExtra") # For producing a nice table 1
library(table1)
```

@@

@@

```{r, warning=FALSE, message=FALSE}
(table1(~ factor(sex) + age + factor(race) + factor(hispanic) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
(table1(~ factor(education) + factor(fam_income) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
(table1(~ factor(employ_status) + factor(paid_sick) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
(table1(~ factor(health_ins_status) + factor(med_access) + factor(med_care_delayed_cost) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
(table1(~ factor(health_status) + factor(bmi) + factor(asthma) + factor(cancer) + factor(chd) + factor(alcohol_status) + factor(smoking_status) | flu_vaccine, data=mydata2))
```

@@

```{r, warning=FALSE, message=FALSE}
(table1(~factor(worry_freq) + factor(dep_freq) | flu_vaccine, data=mydata2))

@@

## Full table for demogrophics characteristics
```{r, warning=FALSE, message=FALSE}
table_dems <- (table1(~ sex + age + race + hispanic + education + fam_income + employ_status + paid_sick + health_ins_status + med_access + med_care_delayed_cost + health_status + bmi + asthma + cancer + alcohol_status + smoking_status| flu_vaccine, data=mydata2))

@@

```{r}

library(table1)

table_dems6 <- table1(~ sex + age + race + hispanic + education + fam_income + employ_status + paid_sick + health_ins_status + med_access + med_care_delayed_cost + health_status + bmi + asthma + cancer + alcohol_status + smoking_status | flu_vaccine, data = mydata2, 
                     label = list(
                       sex = "Sex",
                       age = "Age",
                       race = "Race",
                       hispanic = "Hispanic Ethnicity",
                       education = "Education",
                       fam_income = "Family Income",
                       employ_status = "Employment Status",
                       paid_sick = "Paid Sick Leave",
                       health_ins_status = "Health Insurance Status",
                       med_access = "Medical Access",
                       med_care_delayed_cost = "Delayed Medical Care Cost",
                       health_status = "Health Status",
                       bmi = "Body Mass Index",
                       asthma = "Asthma",
                       cancer = "Cancer",
                       alcohol_status = "Alcohol Consumption",
                       smoking_status = "Smoking Status",
                       flu_vaccine = "Flu Vaccine"
                     ),
                     show_na = TRUE
                    )

table_dems6
```

@@

@@

######## CLEANING TABLE 1 #######

```{r}
 (table1(~ sex + age + race + hispanic + education + fam_income + employ_status + paid_sick + health_ins_status + med_access + med_care_delayed_cost + health_status + bmi + asthma + cancer + chd + alcohol_status + smoking_status| flu_vaccine, data=mydata2))

```{r}
mydata2 %>% tbl_summary(include = c(age, sex))
```

```{r}
 tbl_summary(
    mydata2,
    include = c(sex, age, race, hispanic, education, fam_income, employ_status,    paid_sick, health_ins_status, med_access, med_care_delayed_cost, health_status, bmi, asthma, cancer, alcohol_status, smoking_status),
    by = flu_vaccine, # split table by group
    missing = "no" # don't list missing data separately
  ) %>%
  add_n() %>% # add column with total number of non-missing observations
  add_p() %>% # test for a difference between groups
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels()
```

### Cleaner version of Table 1: Demographic Characteristics
```{r}
table_dems_v3 <- mydata2 %>%
  tbl_summary(
    statistic = list(all_categorical() ~ "{n} ({p}%)", all_continuous() ~ "{mean} ({sd})"),
    by = flu_vaccine,
    missing = "no",
    label = list(
      sex ~ "Sex",
      age ~ "Age",
      race ~ "Race",
      hispanic ~ "Hispanic Ethnicity",
      education ~ "Education",
      fam_income ~ "Family Income",
      employ_status ~ "Employment Status",
      paid_sick ~ "Paid Sick Leave",
      health_ins_status ~ "Health Insurance Status",
      med_access ~ "Medical Access",
      med_care_delayed_cost ~ "Delayed Medical Care Cost",
      health_status ~ "Health Status",
      bmi ~ "Body Mass Index",
      asthma ~ "Asthma",
      cancer ~ "Cancer",
      alcohol_status ~ "Alcohol Consumption",
      smoking_status ~ "Smoking Status"
    )
  ) %>%
  bold_labels() %>%
  add_overall() %>%
  add_n() %>%
  add_p() %>%
  modify_header(label = "**Variable**")

table_dems_v3
```

### 2023-05-05 products/manuscript/Manuscript.qmd
More edits v2

Influenza, or flu, is an extremely contagious respiratory illness caused by a virus. It can cause a range of symptoms, including fever, cough, sore throat, body aches, headache, chills, and fatigue (@cdc2022c.) It can be particularly dangerous for certain groups, such as older adults, young children, and individuals with underlying health issues, often resulting in severe illness or even death. Although influenza vaccinations are widely accessible, many people in the United States do not receive them as recommended. This study examines influenza vaccination uptake among US adults from 2012-2021. Using a subset of the National Health Interview Survey (NHIS) data set, researchers provide important insights into which groups are at highest risk for under-coverage of flu vaccine in order to better inform targeted public health interventions for improving uptake.

@@

Influenza, commonly known as the flu, is a contagious respiratory illness which comes from influenza viruses. It can cause mild to severe illness, and in some cases, it can lead to hospitalization or even death. Symptoms of the flu include fever, cough, sore throat, body
aches, headache, chills, and fatigue (@cdc2022.) It can spread easily from person to person through droplets that are released infected individuals cough, sneeze, or talk.

Vaccines are a crucial tool in preventing the spread of infectious diseases, including the flu. On an individual level, the best way to protect against the flu is by getting vaccinated each year. At the population level, high vaccine coverage is a powerful strategy for reducing spread of the virus (@cdc2022c.)

The U.S. Centers for Disease Control and Prevention (CDC) recommends that everyone six months of age and older receive a flu vaccine each year, with rare exceptions (@cdc2022c.) Influenza is a constantly evolving virus that mutates rapidly, and uncontrolled spread gives rise to many different strains, so new vaccines are developed each year to protect against the most common strains of the virus that are expected to circulate during the upcoming flu season. It is especially important for individuals who are at higher risk of severe illness or complications from the flu, such as young children, older adults, pregnant women, and individuals with underlying medical conditions.

In terms of the availability and cost, there have been significant changes for flu vaccines over the past 50 years (@InfluenzaHistoricTimeline2022.) In the 1960s, flu vaccines were not widely available and were primarily given to people considered to be at high risk for complications. The cost of the vaccine was relatively high and was not covered by most insurance plans. In the 1970s and 1980s, flu vaccines became more widely available and were recommended for more people, including children and healthy adults. The cost of the vaccine began to decrease as more people received it, and insurance coverage for the vaccine became more common. In the 1990s and 2000s, the availability of flu vaccines continued to increase, and the cost of the vaccine became more affordable for a larger proportion of the U.S. population.

@@

Despite the availability and recommendation of influenza vaccines, many people in the United States still do not receive them as recommended. In fact, the CDC reported that only 49.2% of adults aged 18 years and older received a flu vaccine during the 2020-2021 flu season (@cdc2022b.) This low uptake of influenza vaccine highlights the need for understanding the factors that contribute to under-coverage of the vaccine in certain populations.

Racial/ethnic minorities have a long history of reduced vaccination rates relative to white Americans (@khan_etal2018.) Since 2010, flu vaccination coverage has been consistently lower among Black, Hispanic, and AI/AN adults. During the 2021--2022 season, flu vaccination coverage was 54% among White adults, 42% among Black adults, 38% among Hispanic adults, and 41% among AI/AN adults (@cdc2022b.) There are many reasons for these inequities, including lack of access to health care and insurance, missed opportunities to vaccinate, and misinformation and distrust.

In prior studies, lower income levels were associated with decreased odds of influenza vaccine receipt (@gaskinEffectIncomeDisparities2023; @vashist_etal2022), and having health insurance coverage has also been shown to be associated with increased influenza vaccine coverage (@cambou_etal2021; @vashist_etal2022). Influenza vaccination rates were also found to be lower in rural than in urban areas among adults of all age groups and both genders (@jain_etal2022). However, there are still many variables of interest from the NHIS dataset that have yet to be evaluated in detail. Additionally, the periods of study from all prior research all end prior to the onset of the COVID-19 pandemic, which could have altered or heightened disparities even more.

@@

The results of our analysis showed that age was significantly associated with flu vaccine uptake (OR = 1.031, 95% CI = 1.030-1.032, p < 0.001), with older individuals being more likely to have received the vaccine. The sex of the individual was also significantly associated with flu vaccine uptake (OR = 1.416, 95% CI = 1.384-1.448, p < 0.001), with females being more likely to have received the vaccine compared to males. Race was also significantly associated with flu vaccine uptake. Specifically, individuals who identified as black (OR = 0.790, 95% CI = 0.776-0.805, p < 0.001) were less likely to have received the vaccine compared to those who identified as white. Native Americans (OR = 1.284, 95% CI = 1.149-1.436, p = 0.0001) and Asians (OR = 1.141, 95% CI = 1.089-1.195, p < 0.001) were more likely to have received the vaccine than whites. Hispanic individuals (OR = 0.825, 95% CI = 0.806-0.845, p < 0.001) were less likely to have received the vaccine compared to non-Hispanic individuals. Education level was also significantly associated with flu vaccine uptake, with individuals who had higher levels of education being more likely to have received the vaccine. Specifically, individuals with some college education (OR = 1.172, 95% CI = 1.126-1.220, p < 0.001), a bachelor's degree (OR = 1.460, 95% CI = 1.402-1.520, p < 0.001), or a graduate degree (OR = 1.911, 95% CI = 1.818-2.007, p < 0.001) were more likely to have received the vaccine compared to those with no education. Individuals with family incomes of $50,000-$99,999 (OR = 1.140, 95% CI = 1.112-1.168, p < 0.001) and those with family incomes of $100,000 or more (OR = 1.417, 95% CI = 1.380-1.455, p < 0.001) were more likely to have received the vaccine compared to those with family incomes less than $50,000. In terms of health-related variables, individuals with asthma (OR = 1.398, 95% CI = 1.361-1.435, p < 0.001), cancer (OR = 1.322, 95% CI = 1.281-1.364, p < 0.001), and coronary heart disease (OR = 1.637, 95% CI = 1.560-1.717, p < 0.001) were more likely to have received the vaccine compared to those without these conditions. Finally, smoking status was significantly associated with flu vaccine uptake, with former smokers (OR = 1.059, 95% CI = 1.032-1.087, p = 0.0006) being more likely to receive the vaccine compared to nonsmoker, and current smokers (OR = 0.669, 95% CI = 0.643 - 0.696) being less likely to receive vaccine. Finally, former drinkers (OR = 1.059, 95% CI = 1.010-1.111, p = 0.0172) were more likely to receive the vaccine compared to nondrinkers, and this was statistically significant. However, current drinkers (OR = 0.978, 95% CI = 0.941 - 1.016) were less likely to receive vaccine compared to nondrinkers. Although this association was not statistically significant.

>>>>

Influenza, or flu, is an extremely contagious respiratory illness caused
by a virus. It can cause a range of symptoms, including fever, cough,
sore throat, body aches, headache, chills, and fatigue (CDC (@2022c).)
It can be particularly dangerous for certain groups, such as older
adults, young children, and individuals with underlying health issues,
often resulting in severe illness or even death. Although influenza
vaccinations are widely accessible, many people in the United States do
not receive them as recommended. This study examines influenza
vaccination uptake among US adults from 2012-2021. Using a subset of the
National Health Interview Survey (NHIS) data set, researchers provide
important insights into which groups are at highest risk for
under-coverage of flu vaccine in order to better inform targeted public
health interventions for improving uptake.

@@

Influenza, commonly known as the flu, is a contagious respiratory
illness which comes from influenza viruses. It can cause mild to severe
illness, and in some cases, it can lead to hospitalization or even
death. Symptoms of the flu include fever, cough, sore throat, body
aches, headache, chills, and fatigue (@CDC (2022a).) It can spread
easily from person to person through droplets that are released infected
individuals cough, sneeze, or talk.

Vaccines are a crucial tool in preventing the spread of infectious
diseases, including the flu. On an individual level, the best way to
protect against the flu is by getting vaccinated each year. At the
population level, high vaccine coverage is a powerful strategy for
reducing spread of the virus (@CDC (2022c).)

The U.S. Centers for Disease Control and Prevention (@CDC) recommends
that everyone six months of age and older receive a flu vaccine each
year, with rare exceptions (@CDC (2022c).) Influenza is a constantly
evolving virus that mutates rapidly, and uncontrolled spread gives rise
to many different strains, so new vaccines are developed each year to
protect against the most common strains of the virus that are expected
to circulate during the upcoming flu season. It is especially important
for individuals who are at higher risk of severe illness or
complications from the flu, such as young children, older adults,
pregnant women, and individuals with underlying medical conditions.

In terms of the availability and cost, there have been significant
changes for flu vaccines over the past 50 years (@Influenza Historic
Timeline \| Pandemic Influenza (Flu) \| CDC (2022).) In the 1960s, flu
vaccines were not widely available and were primarily given to people
considered to be at high risk for complications. The cost of the vaccine
was relatively high and was not covered by most insurance plans. In the
1970s and 1980s, flu vaccines became more widely available and were
recommended for more people, including children and healthy adults. The
cost of the vaccine began to decrease as more people received it, and
insurance coverage for the vaccine became more common. In the 1990s and
2000s, the availability of flu vaccines continued to increase, and the
cost of the vaccine became more affordable for a larger proportion of
the U.S. population.

@@

Despite the availability and recommendation of influenza vaccines, many
people in the United States still do not receive them as recommended. In
fact, the CDC reported that only 49.2% of adults aged 18 years and older
received a flu vaccine during the 2020-2021 flu season (@CDC (2022b).)
This low uptake of influenza vaccine highlights the need for
understanding the factors that contribute to under-coverage of the
vaccine in certain populations.

Racial and ethnic minorities have a long history of reduced vaccination
rates relative to white Americans (@Khan, @Hall, @Tanner, & @Marlow
(@2018).) Since 2010, flu vaccination coverage has been consistently
lower among Black, Hispanic, and AI/AN adults. During the 2021--2022
season, flu vaccination coverage was 54% among White adults, 42% among
Black adults, 38% among Hispanic adults, and 41% among AI/AN adults
(@CDC (2022b).) There are many reasons for these inequities, including
lack of access to health care and insurance, missed opportunities to
vaccinate, and misinformation and distrust.

In prior studies, lower income levels were associated with decreased
odds of influenza vaccine receipt (@Gaskin, Woods, Ghosh, Watson, &
Huber (2023); Vashist, Choi, & Patel (2022)), and having health
insurance coverage has also been shown to be associated with increased
influenza vaccine coverage (Cambou, Copeland, Nielsen-Saines, & Macinko
(2021); Vashist et al. (2022)). Influenza vaccination rates were also
found to be lower in rural than in urban areas among adults of all age
groups and both genders (Jain et al. (2022)). However, there are still
many variables of interest from the NHIS dataset that have yet to be
evaluated in detail. Additionally, the periods of study from all prior
research all end prior to the onset of the COVID-19 pandemic, which
could have altered or heightened disparities even more.

## Question:

Looking at whether study participants received any influenza vaccine in
the past twelve month, this study aims to answer the following question:

Which groups are at highest risk for low uptake of the influenza
vaccine?

@@

The results of our analysis showed that age was significantly associated
with flu vaccine uptake (OR = 1.031, 95% CI = 1.030-1.032, p \< 0.001),
with older individuals being more likely to have received the vaccine.
The sex of the individual was also significantly associated with flu
vaccine uptake (OR = 1.416, 95% CI = 1.384-1.448, p \< 0.001), with
females being more likely to have received the vaccine compared to
males. Race was also significantly associated with flu vaccine uptake.
Specifically, individuals who identified as black (OR = 0.790, 95% CI =
0.776-0.805, p \< 0.001) were less likely to have received the vaccine
compared to those who identified as white. Native Americans (OR = 1.284,
95% CI = 1.149-1.436, p = 0.0001) and Asians (OR = 1.141, 95% CI =
1.089-1.195, p \< 0.001) were more likely to have received the vaccine
than whites. Hispanic individuals (OR = 0.825, 95% CI = 0.806-0.845, p
\< 0.001) were less likely to have received the vaccine compared to
non-Hispanic individuals. Education level was also significantly
associated with flu vaccine uptake, with individuals who had higher
levels of education being more likely to have received the vaccine.
Specifically, individuals with some college education (OR = 1.172, 95%
CI = 1.126-1.220, p \< 0.001), a bachelor's degree (OR = 1.460, 95% CI =
1.402-1.520, p \< 0.001), or a graduate degree (OR = 1.911, 95% CI =
1.818-2.007, p \< 0.001) were more likely to have received the vaccine
compared to those with no education. Individuals with family incomes of
\$50,000-\$99,999 (OR = 1.140, 95% CI = 1.112-1.168, p \< 0.001) and
those with family incomes of \$100,000 or more (OR = 1.417, 95% CI =
1.380-1.455, p \< 0.001) were more likely to have received the vaccine
compared to those with family incomes less than \$50,000. In terms of
health-related variables, individuals with asthma (OR = 1.398, 95% CI =
1.361-1.435, p \< 0.001), cancer (OR = 1.322, 95% CI = 1.281-1.364, p \<
0.001), and coronary heart disease (OR = 1.637, 95% CI = 1.560-1.717, p
\< 0.001) were more likely to have received the vaccine compared to
those without these conditions. Finally, smoking status was
significantly associated with flu vaccine uptake, with former smokers
(OR = 1.059, 95% CI = 1.032-1.087, p = 0.0006) being more likely to
receive the vaccine compared to nonsmoker, and current smokers (OR =
0.669, 95% CI = 0.643 - 0.696) being less likely to receive vaccine.
Finally, former drinkers (OR = 1.059, 95% CI = 1.010-1.111, p = 0.0172)
were more likely to receive the vaccine compared to nondrinkers, and
this was statistically significant. However, current drinkers (OR =
0.978, 95% CI = 0.941 - 1.016) were less likely to receive vaccine
compared to nondrinkers. Although this association was not statistically
significant.
